Detection of slow slip events using wavelet analysis of GNSS recordings

### 2021-10-30 paper/paper_bssa.tex
last paragraph

\begin{itemize}
\item We use a wavelet-based signal processing method to detect transients in GNSS data, such as slow slip events.
\item There is a good correlation between detections using GNSS data and independent detections using seismic data.
\item The method could be applied in regions where no tremor are detected in conjunction with slow slip events.
\end{itemize}

@@

In many places, tectonic tremor is observed in relation to slow slip and can be used as a proxy to study slow slip events of moderate magnitude where surface deformation is hidden in Global Navigation Satellite System (GNSS) noise. However, in subduction zones where no clear relationship between tremor and slow slip occurrence is observed, these methods cannot be applied, and we need other methods to be able to better detect and quantify slow slip. Wavelets methods such as the Discrete Wavelet Transform (DWT) and the Maximal Overlap Discrete Wavelet Transform (MODWT) are mathematical tools for analyzing time series simultaneously in the time and the frequency domain by observing how weighted differences of a time series vary from one period to the next. In this paper, we use wavelet methods to analyze GNSS time series and seismic recordings of slow slip events in Cascadia. We use detrended GNSS data, apply the MODWT transform and stack the wavelet details over several nearby GNSS stations. As an independent check on the timing of slow slip events, we also compute the cumulative number of tremor in the vicinity of the GNSS stations, detrend this signal, and apply the MODWT transform. In both time series, we can then see simultaneous waveforms whose timing corresponds to the timing of slow slip events. We assume that there is a slow slip event whenever there is a positive peak followed by a negative peak in the wavelet signal. We verify that there is a good correlation between slow slip events detected with only GNSS data, and slow slip events detected with only seismic data for northern Cascadia. The wavelet-based detection method detects well events of magnitude higher than 6 as determined by independent event catalogs (e.g. ~\citep{MIC_2019}).

@@

In our study, we use a similar approach to previous studies with a different reasoning. We only stack signals at nearby GPS stations, assuming that the longitudinal displacement due to the ETS events will then be the same at each of the GPS stations considered. We suppose that some of the noise component is different at each GPS station and will be eliminated by the stacking. Finally, we assume that the noise and the longitudinal displacement due to the ETS events and the secular plate motion have different time scales, so that the wavelet decomposition will act as a bandpass filter to retrieve the displacement signal and highlight the ETS events. We use wavelet methods to analyze GPS and seismic recordings of slow slip events in Cascadia. Our objective is to verify that there is a good correlation between slow slip events detected with only GNSS data, and slow slip events detected with only seismic data. We thus want to demonstrate that the wavelet-based detection method can be applied to detect slow slip events that may currently be obscured using standard methods. \\

@@

We focused our study on northwest Washington State. For the GNSS data, we used the GPS time series provided by the Pacific Northwest Geodetic Array, Central Washington University. These are network solutions in ITRF2008 with phase ambiguities resolved. Solutions are computed with JPL/NASA orbits and satellite clocks. North, East, and Vertical directions are available. However, as the direction of the secular plate motion is close to the East direction, we only used the East direction of the GPS time series for the data analysis, as it has the best signal-to-noise ratio. The wavelet method works best with data with zero mean, and no sharp discontinuities; so we use the cleaned dataset, that is GPS times series with linear trends, steps due to earthquakes or hardware upgrades, and annual and semi-annual sinusoids signals simultaneously estimated and removed following ~\citet{SZE_2004}. For the seismic data, we used the tremor catalog from \textcolor{red}{reference for the catalog. The following is to be modified. the Pacific Northwest Seismic Network (PNSN) ~\citep{WEC_2010}. Tremor were detected and located using waveform envelope correlation and clustering and a centroid location is available for every given five-minute time window when tremor was detected. As the catalog starts in August 2009, we only looked at GPS data recorded in 2009 or later.} \\

@@

We stacked the 8th level detail of the wavelet decomposition of the displacement over all the GPS stations located in a 50 km radius of a given point, for the 16 locations indicated in Figure 3. The result is shown in the top panel of Figure 6, where each line represents one of the locations along strike. To better highlight the peaks in the wavelet details, we highlighted in red the time intervals where the amplitude of the stacked detail is higher than a threshold, and in blue the  time intervals where the amplitude of the stacked detail is lower than minus the threshold. To compare the GPS signal with the tremor signal, we plotted the 8th level detail of the wavelet decomposition of the tremor count on the bottom panel of Figure 6. We multiplied by -1 the cumulative tremor count for the wavelet decomposition in order to be able to match positive peaks with positive peaks and negative peaks with negative peaks. In the tremor catalog from \textcolor{red}{reference?}, there are 17 tremor events with more than 150 hours of tremor recorded. The events are summarized in Table 1. The time of the event is the start date plus half the duration of the event. \\

@@

For the 6th level detail, we see an additional event in the South in Fall 2009 that is present both in the GPS and the tremor data. It may correspond to the northern extent of a big ETS event occurring in Fall 2009 south of the study area (event 19 in the ~\citet{MIC_2019} catalog). There are three small signals in the GPS data in Spring 2012, Fall 2017, and Winter 2020 that are not present in the tremor data, and are probably false detections. To summarize, we assume that true detections are events present in both GPS and tremor time series, and false detections are events present in the GPS but not in the tremor time series. Then, all the 13 events present on the 8th level detail of the wavelet decomposition are true detections and 14 of the 17 events present on the 6th level detail of the wavelet decomposition are true detections. \\

@@

All 14 events that are sub-events of a bigger event are visible at level 5. However, this may be because the bigger events are clearly seen at levels 6 to 8, and also at smaller time scales. The one small event that is not part of a bigger event (Winter 2009) is visible at level 5 of the wavelet decomposition. However, some other events that are not in the catalog of ~\citet{MIC_2019}'s catalog are also visible in late 2007, early 2010, early 2012, and late 2016. Therefore, it is difficult to differentiate between a true detection and a false detection, and to conclude whether the method can indeed detect events of magnitude 5.

@@

In this paper, we develop and test a new approach for detecting transient events in GPS time series, such as slow slip events. We used wavelet methods to analyze GNSS time series and seismic recordings of slow slip events in Cascadia. We used detrended GNSS data, applied the MODWT transform, and stacked the wavelet details over several nearby GNSS stations. As an independent check on the timing of slow slip events, we also computed the cumulative number of tremor in the vicinity of the GNSS stations, detrended this signal, and applied the MODWT transform. In both time series, we could then see simultaneous waveforms whose timing corresponds to the timing of slow slip events. We assumed that there is a slow slip event whenever the wavelet signal gets above a threshold. We verified that there is a good correlation between slow slip events detected with only GNSS data, and slow slip events detected with only seismic data. The wavelet-based detection method detects all events of magnitude higher than 6 as determined by independent event catalogs (e.g. ~\citep{MIC_2019}). We detected signals in the GPS data that could be magnitude 5 events, but it is not easy to differentiate between true detections and false detections.

@@

The GPS recordings used for this analysis can be downloaded from the PANGA website ~\citep{PANGA} \url{http://www.panga.cwu.edu/}. \textcolor{red}{Reference for tremor catalog}. The Python scripts used to analyze the data and make the figures can be found on the first author's Github account \url{https://github.com/ArianeDucellier/slowslip}. Figure 4 was created using GMT ~\citep{WES_1991}.

@@

This work was funded by the grant from the National Science Foundation XXX. A.D. would like to thank Pr Donald Percival for introducing her to wavelet methods during his excellent class taught at University of Washington.

@@

\begin{table}[hbt!]
\caption{Episodic Tremor and Slip events with M > 6 identified by MODWT in both the GPS and the tremor data. The duration and the number of tremor are from the tremor catalog of \textcolor{red}{reference?}. The event number and the magnitude are from the slow slip catalog of ~\citet{MIC_2019}.}
 \centering
 \begin{tabular}{c c c c c}
 \hline
 Time & Duration (days) & Number of tremor (hours) & Event number & Magnitude \\
 \hline
 2007.06 & 28 & 398 & 3 & 6.68 \\
 2008.36 & 25 & 402 & 10 & 6.56 \\
 2009.35 & 24 & 248 & 16 & 6.49 \\
 2010.63 & 29 & 518 & 24 & 6.54 \\
 2011.60 & 37 & 479 & 30 & 6.47 \\
 2012.72 & 37 & 620 & 34 & 6.54 \\
 2013.71 & 27 & 423 & 41 & 6.58 \\
 2014.65 & 15 & 190 & 48 & 6.03 \\
 2014.89 & 38 & 385 & 51 & 6.40 \\
 2016.11 & 43 & 421 & 54 & 6.79 \\
 2017.23 & 19 & 279 & 59 & 6.61 \\
 2018.49 & 22 & 381 & & \\
 2019.23 & 34 & 195 & & \\
 2019.88 & 16 & 205 & & \\
 2020.79 & 26 & 193 & & \\
 2020.86 & 12 & 162 & & \\
 2021.09 & 14 & 230 & & \\
 \hline
 \end{tabular}
 \end{table}

>>>>

\begin{itemize}
\item We use a wavelet-based signal processing method to detect transients in GNSS data, such as slow slip events.
\item There is a good correlation between detections of slow slip using GNSS data and using tremor data.
\item The method could be applied in regions where no tremor are detected in conjunction with slow slip events.
\end{itemize}

@@

In many places, tectonic tremor is observed in relation to slow slip and can be used as a proxy to study slow slip events of moderate magnitude where surface deformation is hidden in Global Navigation Satellite System (GNSS) noise. However, in subduction zones where no clear relationship between tremor and slow slip occurrence is observed, these methods cannot be applied, and we need other methods to be able to better detect and quantify slow slip. Wavelets methods such as the Discrete Wavelet Transform (DWT) and the Maximal Overlap Discrete Wavelet Transform (MODWT) are mathematical tools for analyzing time series simultaneously in the time and the frequency domain by observing how weighted differences of a time series vary from one period to the next. In this paper, we use wavelet methods to analyze GNSS time series and seismic recordings of slow slip events in Cascadia. We use detrended GNSS data, apply the MODWT transform and stack the wavelet details over several nearby GNSS stations. As an independent check on the timing of slow slip events, we also compute the cumulative number of tremor in the vicinity of the GNSS stations, detrend this signal, and apply the MODWT transform. In both time series, we can then see simultaneous waveforms whose timing corresponds to the timing of slow slip events. We assume that there is a slow slip event whenever there is a positive peak followed by a negative peak in the wavelet signal. We verify that there is a good correlation between slow slip events detected with only GNSS data, and slow slip events detected with only tremor data for northern Cascadia. The wavelet-based detection method detects well events of magnitude higher than 6 as determined by independent event catalogs (e.g. ~\citep{MIC_2019}).

@@

In our study, we use a similar approach to previous studies with a different reasoning. We only stack signals at nearby GPS stations, assuming that the longitudinal displacement due to the ETS events will then be the same at each of the GPS stations considered. We suppose that some of the noise component is different at each GPS station and will be eliminated by the stacking. Finally, we assume that the noise and the longitudinal displacement due to the ETS events and the secular plate motion have different time scales, so that the wavelet decomposition will act as a bandpass filter to retrieve the displacement signal and highlight the ETS events. We use wavelet methods to analyze GPS and tremor recordings of slow slip events in Cascadia. Our objective is to verify that there is a good correlation between slow slip events detected with only GNSS data, and slow slip events detected with only tremor data. We thus want to demonstrate that the wavelet-based detection method can be applied to detect slow slip events that may currently be obscured using standard methods. \\

@@

We focused our study on northwest Washington State. For the GNSS data, we used the GPS time series provided by the Pacific Northwest Geodetic Array, Central Washington University. These are network solutions in ITRF2008 with phase ambiguities resolved. Solutions are computed with JPL/NASA orbits and satellite clocks. North, East, and Vertical directions are available. However, as the direction of the secular plate motion is close to the East direction, we only used the East direction of the GPS time series for the data analysis, as it has the best signal-to-noise ratio. The wavelet method works best with data with zero mean, and no sharp discontinuities; so we use the cleaned dataset, that is GPS times series with linear trends, steps due to earthquakes or hardware upgrades, and annual and semi-annual sinusoids signals simultaneously estimated and removed following ~\citet{SZE_2004}. For the tremor data, we used the tremor catalog from the Pacific Northwest Seismic Network (PNSN) ~\citep{WEC_2010}. \\

@@

We stacked the 8th level detail of the wavelet decomposition of the displacement over all the GPS stations located in a 50 km radius of a given point, for the 16 locations indicated in Figure 3. The result is shown in the top panel of Figure 6, where each line represents one of the locations along strike. To better highlight the peaks in the wavelet details, we highlighted in red the time intervals where the amplitude of the stacked detail is higher than a threshold, and in blue the  time intervals where the amplitude of the stacked detail is lower than minus the threshold. To compare the GPS signal with the tremor signal, we plotted the 8th level detail of the wavelet decomposition of the tremor count on the bottom panel of Figure 6. We multiplied by -1 the cumulative tremor count for the wavelet decomposition in order to be able to match positive peaks with positive peaks and negative peaks with negative peaks. In the tremor catalog from the PNSN, there are 17 tremor events with more than 150 hours of tremor recorded. The events are summarized in Table 1. The time of the event is the start date plus half the duration of the event. \\

@@

For the 6th level detail, we see an additional event in the South in Fall 2009 that is present both in the GPS and the tremor data. It may correspond to the northern extent of a big ETS event occurring in Fall 2009 south of the study area (event 19 in the ~\citet{MIC_2019} catalog). There are three small signals in the GPS data in Winter 2012, Fall 2017, and Winter 2020 that are not present in the tremor data, and may be false detections. To summarize, we assume that true detections are events present in both GPS and tremor time series, and false detections are events present in the GPS but not in the tremor time series. Then, all the 13 events present on the 8th level detail of the wavelet decomposition are true detections and 14 of the 17 events present on the 6th level detail of the wavelet decomposition are true detections. \\

@@

All 14 events that are sub-events of a bigger event are visible at level 5. However, this may be because the bigger events are clearly seen at levels 6 to 8, and also at smaller time scales. The one small event that is not part of a bigger event (Winter 2009) is visible at level 5 of the wavelet decomposition. However, some other events that are not in the catalog of ~\citet{MIC_2019}'s catalog are also visible in late 2007, early 2010, early 2012, and late 2016. Therefore, it is difficult to differentiate between a true detection and a false detection, and to conclude whether the method can indeed detect events of magnitude 5. \\

In Figure 9, we see four smaller events that are not in the catalog of ~\citet{MIC_2019}: at about 2007.5, there is a negative peak followed by a positive peak (that is an event in the opposite direction of what would be expected from slow slip), at about 2010.2, 2012.2 and 2020.2, there are positive peaks followed by negative peaks for all the sixteen locations studied in this paper. Looking back at the original GPS data, there is a small increase in the displacement in the eastern direction that lasts about one or two months at about 2007.5. However, the direction of the displacement does not correspond to a slow slip event, and another cause should be found to explain this signal. There is a decrease in displacement that lasts several months at about 2010.2. This transient may correspond to a long duration slow slip event. There is a small decrease in displacement at about 2012.2. Its amplitude is small but the duration and direction correspond to a slow slip event, so this transient could be a very small slow slip event. Finally, there is also a small decrease in displacement at about 2020.2 that is difficult to interpret.

@@

In this paper, we develop and test a new approach for detecting transient events in GPS time series, such as slow slip events. We used wavelet methods to analyze GNSS time series and tremor recordings of slow slip events in Cascadia. We used detrended GNSS data, applied the MODWT transform, and stacked the wavelet details over several nearby GNSS stations. As an independent check on the timing of slow slip events, we also computed the cumulative number of tremor in the vicinity of the GNSS stations, detrended this signal, and applied the MODWT transform. In both time series, we could then see simultaneous waveforms whose timing corresponds to the timing of slow slip events. We assumed that there is a slow slip event whenever the wavelet signal gets above a threshold. We verified that there is a good correlation between slow slip events detected with only GNSS data, and slow slip events detected with only tremor data. The wavelet-based detection method detects all events of magnitude higher than 6 as determined by independent event catalogs (e.g. ~\citep{MIC_2019}). We detected signals in the GPS data that could be magnitude 5 events, but it is not easy to differentiate between true detections and false detections.

@@

The GPS recordings used for this analysis can be downloaded from the PANGA website ~\citep{PANGA} \url{http://www.panga.cwu.edu/}. The Python scripts used to analyze the data and make the figures can be found on the first author's Github account \url{https://github.com/ArianeDucellier/slowslip}. Figure 4 was created using GMT ~\citep{WES_1991}.

@@

This work was funded by the grant from the National Science Foundation EAR-1358512. A.D. would like to thank Professor Donald Percival for introducing her to wavelet methods during his excellent class on Wavelets: Data Analysis, Algorithms and Theory taught at University of Washington.

@@

\begin{table}[hbt!]
\caption{Episodic Tremor and Slip events with M > 6 identified by MODWT in both the GPS and the tremor data. The duration and the number of tremor are from the tremor catalog of the PNSN. The event number and the magnitude are from the slow slip catalog of ~\citet{MIC_2019}.}
 \centering
 \begin{tabular}{c c c c c}
 \hline
 Time & Duration (days) & Number of tremor (hours) & Event number & Magnitude \\
 \hline
 2007.06 & 28 & 398 & 3 & 6.68 \\
 2008.36 & 25 & 402 & 10 & 6.56 \\
 2009.35 & 24 & 248 & 16 & 6.49 \\
 2010.63 & 29 & 518 & 24 & 6.54 \\
 2011.60 & 37 & 479 & 30 & 6.47 \\
 2012.72 & 37 & 620 & 34 & 6.54 \\
 2013.71 & 27 & 423 & 41 & 6.58 \\
 2014.65 & 15 & 190 & 48 & 6.03 \\
 2014.89 & 38 & 385 & 51 & 6.40 \\
 2016.11 & 43 & 421 & 54 & 6.79 \\
 2017.23 & 19 & 279 & 59 & 6.61 \\
 2018.49 & 22 & 381 & & \\
 2019.23 & 34 & 195 & & \\
 2019.88 & 16 & 205 & & \\
 2020.79 & 26 & 193 & & \\
 2020.86 & 12 & 162 & & \\
 2021.09 & 14 & 230 & & \\
 \hline
 \end{tabular}
 \end{table}
